Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-13 article/main.tex
Rajout de l'algo C connu

\subsection{Brute-Force Attack with known $C$}

In the case where $C$ is known, if we find $S_0$, we can calculate every $S_i$
with the given formula, and predict all the following outputs of the
generator. The naive way is by doing a comprehensive research of $upS_0$, and
then calculating $lowS_0$ with the formula :
\begin{equation}
    lowS_0 = upS_0\ \mathtt{XOR}\ \mathtt{unrotate}(X_0)
\end{equation}
with \texttt {unrotate} being the inverse function of \texttt{rotate}.

If our guessing of $upS_0$ is correct, we can obtain the correct $X_1$ with
$S_1$ deduced from the initial $S_0$. The preceding is a naive method for
predicting PCG generator in $2^{64}$ operations. But we can have other values
that verify that test. This will not return $S_0$ but a very short list of
possibles $S_0$, with the actual one in it. We just need to test the following
outputs to find the real $S_0$. This is yet a very slow way to guess the
internal state, so we need a better solution.

\subsection{Brute-Force Attack with unknown $C$}

@@

%\subsection{Different solutions for Truncated Linear Congruential Algorithms}

\subsection{Frieze \textit{et al} algorithm}

The algorithm we studied uses lattices, and more specifically use the\break Lenstra–Lenstra–Lovász lattice basis reduction algorithm (LLL algorithm)\cite{LLLpastack}. A large part of this algorithm is very well explained on Stack Overflow\cite{LLLstack}.
We will detail the entire algorithm here.\\

A lattice in $\mathbb{R}^n$ is a subgroup of the additive group $(\mathbb{R}^n,+)$, isomorphic to $(\mathbb{Z}^n,+)$. The LLL algorithm is an approximation algorithm of the reduction of a lattice basis. It returns a reduced basis, with shorter vectors than the ones given. These vectors are a good approximation of the basis with the smallest determinant.\\

To use this algorithm, we reduce the sequence of $S_i$ to a geometric sequence $S'$ and, given the relationship between $S$ and $S'$, we can retrieve $S$. There are different methods to achieve this result and we selected two of them in our optimised algorithms. Consider the following geometric sequence:
\begin{equation}
    S'_i=\alpha ^i \times S'_0 \mod{M}
\end{equation}
With $S'$ as the vector made of all the $S'_i$ of length $nbiter$, and $A$ blocked, the $S'$ vectors are the vectors of the lattice generated by the matrix :

\begin{equation}
L=
\begin{pmatrix} 
M & 0 & & \dots & 0\\
\alpha & -1 & 0 & \dots & 0\\
\alpha ^2 & 0 & -1 & \dots & 0\\
\vdots & & & \ddots & \vdots\\
\alpha ^{nbiter - 1} & 0 & 0 & \dots & -1
\end{pmatrix}
\end{equation}

%Beau "ell" : $\ell$

Let $Y$ be the most significant bits of the internal states of LCG, and $Z$ be the lower bits, with $S = Y + Z$. We already know $Y$ from the outputs and we are searching for $Z$. 
\begin{itemize}
    \item Let $Y'$ be the upper bits of $S'$;
    \item Let $Z'$ be the lower bits;
    \item Let $L'$  be the reduced matrix returned by the LLL algorithm for $L$ ($L' = LLL(L)$);
\end{itemize}

By definition of $L$, we have :
\begin{align}
    L \times S' &=0 \mod{M}\\
    L' \times S' &= 0 \mod{M}\\
    L' \times (Y' + Z') &= M \times \ell\ \text{($\ell$ unknown integer vector)}\\
    L' \times Z' &= M \times \ell - L' \times Y'
\end{align}
We have $M = 2^k$. L' coefficients are small. We want to find the actual value of $\ell$.\\

We suppose we have every coordinate of $L' \times Z'$ smaller than $M/2$. Then every coordinate of $M \times \ell - L' \times Y'$ is smaller than $M/2$ and $M \times \ell$ is the closest vector of $M$ multiples from $L' \times Z'$. We have $\ell$ is the closest integer vector from $(B \times y)/M$.\\

The only unknown variable left is $Z'$ and :
\begin{equation}
    Z' = L'^{-1} \times (M \times k) - Y' 
\end{equation}
Now we know $Z'$, and the integer $S' \mod{M}$, and finally, $S \mod{M}$ the internal state we were looking for.\\

\subsection{Application to our generator}
The $L$ matrix for our problem is :
\begin{equation}
    L=
\begin{pmatrix}
1208925819614629174706176 & 0 &  0 &  0 &  0 &  0 &  0 &  0 &  0 & 0\\
442210214475561085761093 & -1 &  0 &  0 &  0 & 0 &  0 &  0 &  0 &  0\\
548907270092662740397721 & 0 & -1 &  0 &  0 & 0 &  0 &  0 &  0 &  0\\
68528166612430993102141 & 0 &  0 & -1 &  0 & 0 &  0 &  0 &  0 &  0\\
579003058182980434679665 & 0 &  0 &  0 & -1 & 0 &  0 &  0 &  0 &  0\\
907267876411637163239285 & 0 &  0 &  0 &  0 & -1 &  0 &  0 &  0 &  0\\
971193525585951128490121 & 0 &  0 &  0 &  0 & 0 & -1 &  0 &  0 &  0\\
423043601822819931331309 & 0 &  0 &  0 &  0 & 0 &  0 & -1 &  0 &  0\\
722352710310758831226849 & 0 &  0 &  0 &  0 & 0 &  0 &  0 & -1 &  0\\
239218985093571541740965 & 0 &  0 &  0 &  0 & 0 &  0 &  0 &  0 & -1
\end{pmatrix}
\end{equation}
Each coefficient is about 128 bit long. Here is the $L'$ (using the LLL algorithm of Sagemath \cite{sagemath}):
\begin{equation}
\begin{pmatrix}
    L' =
    77 & -105 & -115 &  17 & -108 & -77  & -7 &  50 & 107 &  17\\
     50  & 80 & 9 & -90 &  138 & -104  & 26  & 90 & -17 & -134\\
    14 & -42 & 130 & -67 & -31 & -172 & 122 & -76 & -32 &  14\\
    -61 &  97 &  70 & 161  & -4 & -229 & -23 & -44  & -1 & -26\\
    118  & 33 & -65 &  -3 &  83 & -176 & -187 & -70 &  25 &  34\\
    111 & -151 & -73 & 115 &  36 & 200 &  79  & -1 & -14 & -82\\
    94  & 15 & -18 & -36 & 142 & -47 &  32 & -188 & 161 & -127\\
    126 & -10 & 129 & 173 & -71  & 61 & -137 &  69 & 7 &  77\\
    174 & -62 &  58  & 39 &  79 & -133 &  93 & 115 &  24 & 181\\
    48 & 240 &  -2 & 114 & -15 &  37 & 145 & -100 &  24 & -19\\
\end{pmatrix}
\end{equation}
This algorithm is clearly really efficient, the coefficients of $L'$ are really small. Its bigger coefficient is $240$. Its smaller than $2^8 = 256$. In our case $M = 2^{64 + known\_low}$. In our tests, we had $known\_low = 16$. We need $Z'$ to be less than 72 bits long, which isn't hard to attain.

\section{lattice geometric progression}

One of the key point of the algorithm is, given a modulus \(M\) and a given common ration \(\alpha\), to find a modular geometric progression knowing only the most significant bits of its \(n\) first terms.

We call \(L(\alpha,M,n)\) the set of n-tuples \(U = (u_0,\dots u_{n-1})\) satisfying :

\[u_{i} = \alpha u_{i-1} \mod M \] for all \(i\) in \(\{1,\dots,n-1\}\).

We easily notice that, if \(U\) and \(V\) are in \(L(\alpha,M,n)\), so are \(-U\) and \(U+V\), hence \(L(\alpha,M,n)\) is a lattice.

The lattice \(L(\alpha,M,n)\) is spawned by the lines of the following matrix :

\[G =\begin{pmatrix}
1&\alpha&\alpha^2&\dots&\alpha^{n-1}\\
0&M&0&\dots&0\\
0&0&M&\dots&0\\
\dots&\dots&\dots&\dots&\dots\\
0&0&0&\dots&M\\
\end{pmatrix}\]

We want to find an element \(U\) in the lattice only knowing the most significant bits of each of its coordinates. We decompose \(U = U_1 + U_2\) where \(U_1\) contains only the upper bits and \(U2\) the lowers one. There is no reason for \(U_1\) to be in the lattice, but as \(U_2\) is small, we know \(U_1\) is \emph{close} to \(U\). So, to find \(U\), we will solve CVP(\(L(\alpha,M)\),\(U_1\)).

\section{Predicting PCG with Known Increment}
\subsection{Reduction to a truncated LCG model}
By exploiting the nature of this generator, it becomes apparent that some values can be easily guessed. We already saw that if we guess $rot_i$, then retrieving $X_i$ becomes a simple \texttt{XOR} operation, and in turn, if we have some knowledge of a particular bit of $lowS_i$ (respectively $upS_i$), we can deduce the equivalent bit from $upS_i$ (respectively $lowS_i$).\\

Moreover, there is an interesting proprity concerning the least significant bits of the internal state. Indeed, if we define $lowC$ as the 64 least significant bits of $C$ ($lowC = C \mod{2^{k/2}})$, we have:

\begin{align}
    S_{i+1} &= A \times S_i + C \mod{2^k}\\
    upS_{i+1} \times 2^{k/2} + lowS_{i+1} &= A \times upS_i \times 2^{k/2} + A \times lowS_i + C \mod{2^k} \\
    lowS_{i+1} &= A \times lowS_i + lowC \mod 2^{k/2}
\end{align}
We can see that $lowS_{i+1}$ is independent from $upS_i$. That mean that knowing $lowS_i$, we can deduce all $lowS_j$ with $j>i$. This relationship is used in the Brute-Force attack with known $C$, in the previous section.
We can see on (Figure \ref{pcg128goodbits})%figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.70\linewidth]{pictures/deducingPCG128.png}
    \caption{PCG128 : Relevant bits of the outputs}
    \label{pcg128goodbits}
\end{figure}
that we can extract a lot of information by brute forcing certain well chosen bits. We defined $known\_up$ (respectively $known\_low$) the number of upper (respectively lower) bits searched by brute force.

\subsection{Reduction to a geometric sequence and validation of our values}

$C$ is known. We are looking for a geometric sequence $S'$ with which we can validate our brute-forced bits and extract $S_0$. Let $polC$ be the vector where $polC_i = \sum_{j=0}^{i} C \times A^j$ and $S'$ where $S'_i = S_i - polC_i$. We have :
\begin{align}
    S'_{i+1} &= S_{i+1} - polC_{i+1}\\
    S'_{i+1} &= A \times S_i + C - polC_{i+1}\\
    S'_{i+1} &= A \times S_i + C - (C + A \times C + ...+ A^{i} \times C)\\
    S'_{i+1} &= A \times (S'_i + polC_i) - A \times (C + ... + A^{i-1} \times C)\\
    S'_{i+1} &= A \times S'_i + A \times polC_i - A \times polC_i\\
    S'_{i+1} &= A \times S'_i
\end{align}

We have a convenient $S'$ that can be used in the algorithm presented in the previous section and, there exist a simple bijection between $S'$ and $S$. However, its important to notice that $S'$ will not be computed modulo $M = 2^k$ but modulo $N = 2^{k/2 + known\_low}$, because we are doing an comprehensive research of the middle bits of $S$ and not the upper ones (except for the 6 bits of $rot$). This is not a problem, because when we have the values of $X$, and $lowS$, we can always obtain $upS$ and ultimately $S$.\\

Using this bijection, we can validate if our chosen values are correct. Indeed, after finding $Z'$ through the LLL algorithm, we can compute $S'_0$ and $S'_1$ modulo $2^{k/2 + known\_low}$, then $S_0$ and $S_1$ modulo $2^{k/2 + known\_low}$, and finally $S_1$ and $S_0$. We verify if $S_1$ is coherent with $S_0$. If it is, then we have guessed the correct values for our generator, if not, we then consider this another value as our next test.

\newpage

\subsection{Pseudo-Algorithm}

\begin{algorithm}[h!]
\caption{Algorithm with known $C$}
\begin{algorithmic}
\REQUIRE $X$ the list of outputs
\STATE computation of the $LLL$ matrix
\STATE computation of $polC$
\FOR{ all $rot_i$ for $i$ in range($nbiter$) }
    \FOR{i in range($nbiter$)}
        \STATE \texttt{unrotate}($X_i$)
    \ENDFOR
    \FOR{all $W_0$}
        \STATE calculate the other $W_i$ with $W_0$ and $C$
        \STATE calculate $Y$ using $W_i\ \mathtt{XOR}\ X_i$ and $rot_i\ \mathtt{XOR}\ X_i$
        \STATE $W' = W - polC$ (we subtract only the corresponding bits of $C$)
        \STATE $Y' = Y - polC$
        \STATE calculation of $Z'$ due to LLL algorithm
        \IF{$2^{k/2-known\_up}>Y'>=2^{known\_low}$}
            \IF{$Z$ is correct}
                \STATE return the corresponding $S_0$
            \ENDIF
        \ENDIF
    \ENDFOR
\ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\subsection{Performance test results}
Our code is implemented in Sagemath. This choice was made because of the development advantages it offered: to be able to work with arbitrarily large integers and have access to a built-in LLL function.\\

The number of bits obtained by exhaustive research, as we have seen in the first section, is $nbiter \times known\_up + known\_low$. The most efficient way should be to only guess the lower bits ($known\_up = 0$). But we need the 6 bit value of $rot_i$ in our algorithm. So we kept $known\_up = 6$ and searched for the best equilibrium between $nbiter$ and $known\_low$. \\

Moreover, we also need this algorithm to be reliable. That is a concern because we know that there is a rounding error caused by an Euclidian Division. Depending on the error, it is possible that the algorithm doesn't find the correct $Z$ even if $Y$ and $W$ are correct (with $W$ the lower bits of $S$ and $Y$ the middle bits of $S$ found by \texttt{XOR}). \\

We tested the reliability and the efficiency of the algorithm. By trial and error, we ended up finding two interesting settings :
\begin{itemize}
    \item $known\_low = 16$ and $nbiter = 4$ : we have a $2^{40}$ brute-force algorithm. When supplying it with PCG128 outputs, the rate at which the algorithm detects it and obtains the correct $S_0$ is of 90\%.
    \item $known\_low = 15$ and $nbiter = 4$ : we have a $2^{39}$ brute-force algorithm. When supplying it with PCG128 outputs, the rate at which the algorithm detects it and obtains the correct $S_0$ is of 70\%.
\end{itemize}

In any case, if we supply a random output, the algorithm will detect this and will never return a false $S_0$. When not finding $S_0$, the default answer is to disqualify the input as a PCG data steam.\\

Unfortunately, we didn't get to run our entire solution in a real time environment, because with our personal equipment, and not fully optimized code, we estimated that it would have take more or less ten years to obtain an answer, which is obviously too long.
The local environment at the university didn't provide the right tools (SageMath), which complexified the development process of sharing the computation load between multiple computers. However, there are some methods and coding improvements to reduce the time needed, as we will discuss in the next section.

>>>>

@@

\section{C connu}

\subsection{Réduction}
Pour deviner l'état interne du générateur, une méthode consiste à procéder par brute-force. On devine les 64 bits supérieurs de $S_0$, on en déduit la rotation $r_0$ utilisée, on peut retrouver  les 64 bits inférieurs:

\begin{equation}
   S_0[0:64] = S_0[64:128] XOR \texttt{ROT(X_0 , 64 - r_0)} 
\end{equation}

Cette méthode demanderait de l'ordre de 2^64 calculs, ce qui n'est pas acceptable. Nous pouvons cependant utiliser cette équation pour déduire des informations sur l'état interne de S. En effet,en devinant par brute-force les 6 bits du haut de S, et ses \ell bits du bas, on récupère :
% schéma "relevant bits of the outputs"

On réduit $S$ à une série géométrique plus simple, . Notons $S'$ l'état interne auquel on retire sa composante en $C$, et en $S_0[0:\ell]$ (note de bas de page : le fait de retirer $S_0[0:\ell]$ n'est pas nécéssaire, mais nous permet de limiter la taille des instances utilisées à 64 bits, très utile pour la programmation en C). Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, de matrice génératrice :
\begin{equation}
\begin{pmatrix} 
1 & \alpha & \alpha ^2 & \dots & \alpha ^{nbiter - 1}\\
0 & M & 0 & \dots & 0\\
0 & 0 & M & \dots & 0\\
\vdots & & & \ddots & \vdots\\
0 & 0 & \dots & 0 & M
\end{pmatrix}
\end{equation}
On

\subsection{Algorithme}
On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau blablabla par l'algorithme LLL.
Pour tous les $r_i$ et les $S_0[0 : \ell]$ possibles:\\
\begin{enumerate}
\item On récupère le $T$ correspondant à nos suppositions. On calcule $T'$. ($T'_i = (S_i[64 - 6 : 64 + \ell] - (a^i*S_0[0 : \ell] + C * \sum_{j = 0}^i a^j)[64 - 6 : 64 + \ell]) * 2^{64 - 6 - \ell}$)\\
%schema T et T'

\item We get the correspondant $S'[\ell : 64 + \ell]$ by using the babai rounding technique on $T'$ using $G$ and $G^{-1}$.

\item We test our solution by computing $S_0$ and testing if the following internals states are coherent with the PCG outputs. If the results are coherents, we return S_0.

\end{enumerate}

@@
